---
title: "Drug Overdose Analysis"
output:
  pdf_document: default
  html_notebook: default
---



# Preparation


## Import packages

The first step of the analysis is to import the necessary packages:

```{r}
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("janitor")
#install.packages("lubridate")


library("dplyr")
library("ggplot2")
library("stringr")
library("janitor")
library("lubridate")
```

- **dplyr**: provides a set of tools for efficiently manipulating datasets in R.
- **ggplot2**: for data visualization.
- **stringr**: provides tools to convert the case of strings.
- **janitor**: for examining and cleaning dirty data.
- **lubridate**: to simplify working with dates.


## Import data set

The data set was taken from <https://www.kaggle.com/datasets/ruchi798/drug-overdose-deaths>,
which contains information regarding accidental drug related deaths in the U.S
state of Connecticut, in the years between 2012-2018.
The dataset contains a total of 5105 records, and 42 columns.

```{r}
data <- read.csv("drug_deaths.csv")
```

To check that the data was imported correctly, the `head` function can be used.

```{r}
head(data)
```
The first question that be asked about the dataset is what information it contains,
i.e. what are the columns of the dataset.
This question can be answered using the `names` function:

```{r}
names(data)
```

The dataset does not include a description for the columns, but most of them are
quite self-explaining. A few, however, deserve an explanation:

- *X*: a progressive integer to identify the record, starting from 0.
- *ID*: probably used to identify the person without revealing private information.
        It's composed by a two digit integer that represents the year of death,
        and a progressive four-digit integer that starts from zero.
        The two numbers are separated by a "-".
- *DateType*: specifies if the `date` should be interpreted as the date of death
               or the date when the death was reported.
- *Location*: location of death.
- *LocationifOther*: only used if `location=="other"`, to better specify the
                      location of death.
- *COD*: cause of death.
- *OtherSignificant*: other significant factors related to the death.



# Cleaning


## Names

The `janitor` package provides the `clean_names()` function which handles
problematic variable names, by doing the following cleaning steps:
- Return names with only lowercase letters, with _ as a separator
- Handles special characters and spaces
- Appends numbers to duplicated names
- Converts “%” to “percent” to retain meaning

```{r}
data <- data %>% clean_names()
```

The updated names can be seen using the `names` function:
```{r}
names(data)
```


## Unused columns

The dataset contains, for each record, two ids: `x` and `id`.
Neither of them are actually useful for this report, so they can be
removed without any consequence.

There could however be multiple records with the same id, which could
be caused by an error in how the dataset was built.
To verify this, we can use the function `get_dupes()` provided by `janitor`:

```{r}
data %>% get_dupes(x)
data %>% get_dupes(id)
```

Since there are no dupes, the columns can be safely deleted:

```{r}
data <- select(data,-c(x,id))
```

Also, the column `other_significan` is mostly empty:

```{r}
mean(data$other_significan=="")
```

Because it won't be used in this analysis, this column can also be deleted:

```{r}
data <- select(data, -other_significan)
```


## Inconsistent casing

The values in the column `mannerof_death` have inconsistent casing:

```{r}
table(data$mannerof_death)
```

This problem may also be present in other columns of the dataset.
To fix this, we can use the function `str_to_title` provided by the package `stringr`:

```{r}
data <- data %>% mutate_all(str_to_title)
```


## Dates

The dataset contains the date of death as a string, which consists in the 
actual date, plus the time of death. So it can be useful to convert the 
string to an actual date-time class.
However, looking at the data, it can be noticed that for most of the date-times,
the timepart is always equal to "12:00:00 AM". This can be verified:

```{r}
datetimes <- as.POSIXct(data$date, format="%m/%d/%Y %T")
times <- strftime(datetimes, format="%H:%M:%S")
print(unique(times))
```

All of the times of death are either "12:00:00 AM" or NA, so they can be eliminated
from the dataset.
This means that the significant part of the string is just the date:

```{r}
data$date <- as.Date(data$date, "%m/%d/%Y")
```



# Data exploration


## Distribution of deaths


### Over the years

An interesting information that can be easily discovered is the distribution
of the deaths over the years.

```{r}
deaths_per_year <- data %>%
  filter(!is.na(date)) %>%
  group_by(year=format(date,"%Y")) %>%
  summarise(count=n()) 
  

ggplot(data=deaths_per_year,aes(x=year,y=count,label=count,fill=count)) +
  geom_col() +
  labs(title="Deaths per year",x="Year",y="Number of deaths") +
  theme_minimal() +
  theme(legend.position = "none")+
  geom_text(aes(label = count), nudge_y = 40) 
```


### Over the months

```{r}
deaths_per_month <- data %>%
  filter(!is.na(date)) %>%
  group_by(month=month(date,label=TRUE)) %>%
  summarise(count=n()) 
  

ggplot(data=deaths_per_month,aes(x=month,y=count,label=count,fill=count)) +
  geom_col() +
  labs(title="Deaths per year",x="Month",y="Number of deaths") +
  scale_x_discrete(labels=format(ISOdate(2004,1:12,1),"%B")) +
  theme_minimal() +
  theme(legend.position = "none",axis.text.x=element_text(angle=45))+
  geom_text(aes(label = count),nudge_y = 20)
```


### Over the days of week

```{r}
deaths_per_day <- data %>%
  filter(!is.na(date)) %>%
  group_by(day=wday(date,label=TRUE)) %>%
  summarise(count=n()) 

ggplot(data=deaths_per_day,aes(x=day,y=count,label=count,fill=count)) +
  geom_col() +
  labs(title="Deaths per year",x="Day of week",y="Number of deaths") +
  theme_minimal() +
  theme(legend.position = "none",axis.text.x=element_text(angle=45))+
  geom_text(aes(label = count),nudge_y = 20)
```