---
title: "Drug Overdose Analysis"
output:
  pdf_document: default
  html_notebook: default
---

# Preparation

## Import packages

The initial step of the analysis is to import the necessary packages:
```{r import-packages}
#install.packages(c(
#  "tidyverse",
#  "tidytext",
#  "janitor",
#  "lubridate",
#  "usmap",
#  "maptools"
#))

library("tidyverse")
library("tidytext")
library("janitor")
library("lubridate")
library("usmap")
```
- **tidyverse**: a collection of packages for data science, which includes
  dplyr, ggplot2, and others.
- **tidytext**: text mining utilities.
- **janitor**: for examining and cleaning dirty data.
- **lubridadata te**: to make working with dates easier.
- **usmap**: for US map data frames.

## Import data set

The data set was taken from
<https://www.kaggle.com/datasets/ruchi798/drug-overdose-deaths>,
and it contains information regarding accidental drug related deaths in the U.S
state of Connecticut, in the years between 2012-2018.
```{r read-csv}
deaths <- read.csv("drug_deaths.csv")
```
To check that the data was imported correctly, the `head` function can be used.
```{r head-dataset}
head(deaths)
```
The first question that can be asked about the data set is what information it contains,
i.e. what are its columns.
This question can be answered using the `names` function:
```{r names-dataset}
names(deaths)
```
Using the functions `nrow` and `ncol` we can get the number of rows and columns
in the data set.
```{r dimensions-dataset}
nrow(deaths)
ncol(deaths)
```
The data set does not include a description for the columns,
but thankfully most of them are self-explaining.
A few, however, require an interpretation:

- *X*: a progressive integer to identify the record, starting from 0.
- *ID*: probably used to identify the person without revealing private information.
        It's composed by a two digit integer that represents the year of death,
        and a progressive four-digit integer that starts from zero.
        The two numbers are separated by a "-".
- *DateType*: specifies if the `date` should be interpreted as the date of death
               or the date when the death was reported.
- *Location*: location of death.
- *LocationifOther*: only used if `location=="other"`, to better specify the
                      location of death.
- *COD*: cause of death.
- *OtherSignificant*: other significant factors related to the death.

# Cleaning

Now that the data has been imported, the data needs to be cleaned
to standardize the records and prevent inconsistencies.

## Names

The `janitor` package provides the `clean_names()` function which handles
problematic variable names, by doing the following cleaning steps:
- Return names with only lowercase letters, with _ as a separator;
- Handles special characters and spaces;
- Appends numbers to duplicated names;
- Converts “%” to “percent” to retain meaning.
```{r clean-names}
deaths <- deaths %>% clean_names()
```
The updated names can be seen using the `names` function:
```{r cleaned-names}
names(deaths)
```

## Column modifications

The data set contains two ids for each record: `x` and `id`.
This means that one of them can be removed without consequences.

There could however be multiple records with the same id, which could
be caused by an error in how the data set was built.
This can be verified by computing the number of unique values of `x` in the set,
which should be equal to the number of records.
```{r unique-x}
length(unique(deaths$x)) == nrow(deaths)
```
Since there are no duplicates, the `x` column can be safely deleted:
```{r delete-x}
deaths <- deaths %>% select(-x)
```
Another column that won't be used is `DateType`:
```{r delete-date-type}
deaths <- deaths %>% select(-date_type)
```
The data set contains a column named `morphine_not_heroin`,
probably to specify that the person specifically used morphine
instead of heroin, which is produced from morphine.
This column can be renamed to just `heroin`.
```{r rename-morphine}
deaths <- deaths %>% rename(morphine = morphine_not_heroin)
```
There also two columns related to fentanyl: `fentanyl` and `fentanyl_analogue`.
For this analysis they can be reduced to just one column.
```{r reduce-fentanyl}
deaths <- deaths %>%
  mutate(fentanyl = ifelse(fentanyl_analogue == 1, "1", fentanyl))

deaths <- deaths %>% select(-fentanyl_analogue)
```

## Convert values

There could be empty strings in some cells of the data set.
It's useful to convert such empty strings to NA, so they can be easily
excluded from queries.
```{r mutate-all-na}
deaths <- deaths %>% mutate_all(na_if, "")
```
The `sex` field seems to contain a few fields with the `Unknown` value,
which can also be converted to `NA`.
```{r sex-unknown-na}
table(deaths$sex)

deaths <- deaths %>%
  mutate(sex = ifelse(sex == "Unknown", NA, sex))
```
Another couple of columns related to drugs contain weird values:
```{r morphine-fentanyl-weird-values}
deaths %>%
  select(fentanyl, morphine) %>%
  filter((fentanyl != "1" & fentanyl != "0") |
           (morphine != "1" & morphine != "0"))
```
The ones related to fentanyl seem to suggest that the drug was detected,
so they can all be put to `1`.
The cells about morphine instead are unclear so they should be set to NA.
```{r morphine-fentanyl-weird-values-fix}
deaths <- deaths %>%
  mutate(fentanyl = ifelse((fentanyl != "1" &
                              fentanyl != "0"), 1, fentanyl))

deaths <- deaths %>%
  mutate(morphine = ifelse((morphine != "1" &
                              morphine != "0"), NA, morphine))
```

## Inconsistent letter casing

The values in the column `mannerof_death` have inconsistent casing:
```{r inconsistent-casing}
table(deaths$mannerof_death)
```
This problem may also be present in other columns of the data set.
The function `str_to_title` provided by the package `stringr` can be used
to fix this:
```{r inconsistent-casing-fix}
deaths <- deaths %>% mutate_all(str_to_title)
```

## Dates

The data set contains the date of death as a string composed by the
actual date, plus the time of death.

It can be useful to convert the string to an actual date-time class.
However, in the data, for most of the date-times,
the substring that specifies the time is always equal to "12:00:00 AM".
```{r time-always-same}
datetimes <- as.POSIXct(deaths$date, format = "%m/%d/%Y %T")
times <- strftime(datetimes, format = "%H:%M:%S")
print(unique(times))
```
All of the times of death are either "12:00:00 AM" or NA, so they can be eliminated
from the data set.
This means that the significant part of the string is just the date:
```{r set-date}
deaths$date <- as.Date(deaths$date, "%m/%d/%Y")
```

## NA values

```{r plot-na}
missing <- colMeans(is.na(deaths)) * 100
missing <- missing[missing >= 1]
missing <- data.frame(missing)
missing <- rownames_to_column(missing, var = "row_name")

missing$label = sprintf("%0.1f%%", missing$missing)

ggplot(data = missing, aes(
  x = row_name,
  y = missing,
  label = label,
  fill = missing
)) +
  geom_bar(stat = "identity") +
  geom_text(nudge_y = 5.5) +
  coord_flip() +
  scale_x_reordered() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Percentage of NA values for each column", x = "Column", y =
         "Percentage of NA values",)
```
The columns that contain mostly NAs can be removed from the data frame.
```{r delete-na-columns}
deaths <- deaths %>%
  select(-c(
    other_significan,
    other,
    locationif_other,
    injury_state,
    injury_county
  ))
```


# Analysis of the use of drugs

To simplify the extraction of information about the drug abuse, a new data frame
can be built using `pivot_longer`.

This function allows, for each record (which in the original data frame corresponds
to a single person), to create *n* records in the new data frame, where *n* is
the total number of different drugs reported in the data set (heroin, fentanyl,
morphine, ...). Each of these new *n* records will correspond to a single person,
i.e. they will all have the same `id`.
In each of these new rows, a single drug is associated
to a value of `1` if the drug was reported in the original record,
or `0` otherwise.

In other words, `pivot_longer` lengthens the data, increasing the number of rows
and decreasing the number of columns.
```{r deaths-drugs}
deaths_drugs <- deaths %>%
  select(
    "id",
    "sex",
    "age",
    "location",
    "heroin",
    "fentanyl",
    "morphine",
    "cocaine",
    "oxycodone",
    "oxymorphone",
    "ethanol",
    "hydrocodone",
    "benzodiazepine",
    "methadone",
    "amphet",
    "tramad",
    "hydromorphone"
  ) %>% 
  pivot_longer(
    cols = c(
      "heroin",
      "fentanyl",
      "cocaine",
      "oxycodone",
      "oxymorphone",
      "morphine",
      "ethanol",
      "hydrocodone",
      "benzodiazepine",
      "methadone",
      "amphet",
      "tramad",
      "hydromorphone"
    ),
    names_to = "drug_name",
    values_to = "value",
    values_drop_na = TRUE,
  ) %>%
  mutate(value = as.numeric(value))
```

## Most popular drugs

Using this new data frame, the distribution in the use of
drugs can be easily discovered:
```{r most-popular-drugs}
deaths_drugs %>%
  filter(value != 0) %>%
  group_by(drug_name) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = drug_name,
    y = count,
    label = count,
    fill = count
  )) +
  geom_col() +
  labs(title = "Drugs found in subjects", x = "Drug",
       y = "Number of subjects") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  geom_text(nudge_y = 80)  
```
Heroin and fentanyl, both of which are opioids, are the two most
used drugs by a significant margin.
This could be related to the current opioid epidemic in the USA.

Cocaine is also widespread, along with alcohol consumption.

## Number of drugs used

We can also analyze the number of (distinct) drugs used by the subjects.
```{r number-of-drugs}
deaths_drugs %>%
  group_by(id) %>%
  summarise(count = sum(value)) %>%
  count(count) %>%
  ggplot(aes(
    x = count,
    y = n,
    label = n,
    fill = n
  )) +
  geom_col() +
  labs(title = "Distinct drugs found in subjects",
       x = "Number of drugs", y = "Subjects") +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(nudge_y = 60)
```
The fact that there are subject with no drugs reported is noteworthy:
this could suggest an error in the data insertion, or maybe there were
signs of drug overdose on the patient,
but the all the test executed were negative.

## Multiple drugs

We can extract the overall percentage of individuals
that overdosed on multiple drugs:
```{r multiple-drugs}
death_by_multiple_drugs <- deaths_drugs %>%
  group_by(id) %>%
  summarise(count = sum(value)) %>%
  count(count) %>%
  filter(count >= 2) %>%
  reduce(sum)


sprintf("%0.1f%%", death_by_multiple_drugs / count(deaths) * 100)
```

## Combination of opioids and benzodiazepines

From the previous plot we can understand that the majority of individuals
overdosed on two distinct drugs at the same.
This is consistent with a study published in 2019, that associates
the combination of drugs to an increased risk of overdose.

The same study suggest that a particularly lethal combination are opioids
and benzodiazepines. We can verify how many subjects used these two drugs
at the same time:
```{r opioids-benzodazepines}
death_op_ben <- deaths %>%
  filter(any_opioid == "1", benzodiazepine == "1") %>%
  count()

sprintf("%0.1f%%", death_op_ben / count(deaths) * 100)
```
## Single drug

Still, a lot of people overdosed on a single drug. It's interesting to discover
which one it is.
```{r single-drug}
ids <- (deaths_drugs %>%
          group_by(id) %>%
          summarise(count = sum(value)) %>%
          filter(count == 1))[["id"]]

deaths_drugs %>%
  filter(id %in% ids, value == 1) %>%
  group_by(drug_name) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = drug_name,
    y = count,
    label = count,
    fill = count
  )) +
  geom_col() +
  labs(title = "Overdose deaths by single drug", x = "Drug", y = "Number of deaths") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  geom_text(nudge_y = 40)  
```
So, heroin is the drug that is most overdosed on by itself
with fentanyl coming second.
This is not surprising because, as we have seen before,
these two are the two overall most used drugs.

A fascinating result is that ethanol and benzodiazepines, while also
being among the drugs most commonly found in the subjects,
did not cause many overdose victims by themselves.

## Drug usage with alcohol

Another interesting statistic is the number of individuals that used drugs
along with alcohol (ethanol): many studies have confirmed that combining alcohol
with any kind of drug is dangerous.
```{r drugs-with-alcohol}
ids <- (deaths_drugs %>%
          filter(drug_name == "ethanol", value == 1))[["id"]]

used_alcohol <- deaths_drugs %>%
  filter(id %in% ids) %>%
  group_by(id) %>%
  summarise(count = sum(value)) %>%
  count(count) %>%
  filter(count >= 2) %>%
  reduce(sum)

sprintf("%0.1f%%", used_alcohol / count(deaths) * 100)
```
# Analysis of death locations

## TODO: Death location distribution

## Death location for each drug

```{r death-location-drug}
deaths_drugs %>%
  mutate(location = fct_lump_n(location, n = 3)) %>%
  select(location, value, drug_name) %>%
  filter(
    location != "Unknown",
    value != 0,
    drug_name %in% c(
      "heroin",
      "fentanyl",
      "cocaine",
      "benzodiazepine",
      "ethanol",
      "oxycodone"
    )
  ) %>%
  group_by(drug_name, location) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  ggplot(aes(
    x = location,
    y = value,
    label = value,
    fill = location
  )) +
  geom_col(width = 0.8) +
  geom_text(vjust = -0.3, size = 4) +
  scale_y_continuous(expand = expansion(add = c(0, 300))) +
  facet_wrap(~ drug_name, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Death Location by Drug Name",
       x = "Location",
       y = "Count")
```

# Analysis by sex

The data set contains information about both female and male subjects.
To better understand the spread of drug abuse, it is interesting to
discriminate between the two sexes in the analysis.

## Distribution of overdose deaths by sex

We can begin by finding out how many subjects are male and how many are female.
```{r deaths-sex}
deaths %>%
  filter(!is.na(sex), sex != "Unknown") %>%
  group_by(sex) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = sex,
    y = count,
    label = count,
    fill = count
  )) +
  geom_bar(stat = "identity") +
  labs(title = "Overdose deaths distributed by sex",
       x = "Sex", y = "Number of deaths") +
  geom_text(nudge_y = 200,  size = 5) +
  theme_minimal() +
  theme(legend.position = "none")
```

Almost three quarters of all the subjects in the data set are male.

## Most abused drugs for each sex

We can also plot how the abuse of drugs is distributed among the sexes.
```{r drugs-sex}
deaths_drugs %>%
  filter(!is.na(sex), value != 0) %>%
  group_by(drug_name, sex) %>%
  summarise(count = sum(value), .groups = "drop") %>%
  mutate(sex = as.factor(sex),
         drug_name = reorder_within(drug_name, count, sex)) %>%
  ggplot(aes(
    x = fct_reorder(drug_name, count),
    y = count,
    label = count,
    fill = drug_name
  )) +
  geom_col(width = 0.8) +
  scale_y_continuous(expand = expansion(add = c(0, 600))) +
  geom_text(hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_x_reordered() +
  facet_wrap(vars(sex), scales = "free") +
  labs(title = "Most popular drugs for each sex",
       x = "Drug name",
       y = "Number of subjects") +
  theme_minimal() +
  theme(legend.position = "none")
```
Heroin and fentanyl are the two most abused drugs by both sexes.
More females than males abused benzodiazepines, which is the third
most used drug by females and the sixth most popular between males.
Also males seem to have a tendency to consume more alcohol than female subjects.

## TODO: at least one more

# Distribution of overdose deaths over the years

An interesting information that can be easily discovered is the distribution
of the deaths over the years.
```{r deaths-years}
deaths %>%
  filter(!is.na(date)) %>%
  group_by(year = format(date, "%Y")) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = year,
    y = count,
    label = count,
    fill = count
  )) +
  geom_col() +
  labs(title = "Deaths per year", x = "Year", y = "Number of deaths") +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(nudge_y = 40) 
```
In this plot we can see that the number of overdose is skyrocketing each
following year. Once again this is probably related to the increasing
number of drug users addicted to opiods.

## Over the months

To plot the distribution of overdose deaths over the months:
```{r deaths-months}
deaths %>%
  filter(!is.na(date)) %>%
  group_by(month = month(date, label = TRUE)) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = month,
    y = count,
    label = count,
    fill = count
  )) +
  geom_col() +
  labs(title = "Deaths per month", x = "Month", y = "Number of deaths") +
  scale_x_discrete(labels = format(ISOdate(2004, 1:12, 1), "%B")) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  geom_text(nudge_y = 20)
```
This plot does not contain any particularly useful information:
the number of overdose cases is quite consistent. The slight fluctuations
could be caused by the availability of drugs.

## Over the days of week

Finally, we can plot the number of deaths for each day of the week.
```{r deaths-day}
deaths %>%
  filter(!is.na(date)) %>%
  group_by(day = wday(date, label = TRUE)) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = day,
    y = count,
    label = count,
    fill = count
  )) +
  geom_col() +
  labs(title = "Deaths per day of week", x = "Day of week", y = "Number of deaths") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  geom_text(nudge_y = 40)
```
The number of deaths is higher during the weekends: this could be caused
by parties and increased nightlife, where it is easier to come in contact
with drugs.

# Age

```{r age-groups}
deaths_age <- deaths %>%
  select(id,age) %>%
  filter(!is.na(age)) %>% 
  mutate(
    age_group = case_when(
      age < 20             ~ "< 20",
      age >= 20 & age < 30 ~ "20-29",
      age >= 30 & age < 40 ~ "30-39",
      age >= 40 & age < 50 ~ "40-49",
      age >= 50 & age < 60 ~ "50-59",
      age >= 60            ~ "> 60",
    ),
    age_group = factor(
      age_group,
      levels = c("< 20", "20-29", "30-39", "40-49", "50-59", "> 60")
    )
  )
```

```{r deaths-age}
deaths_age %>%
  group_by(age_group) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = age_group,
    y = count,
    fill = age_group,
    label = count
  )) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_text(hjust = -0.2) +
  theme_minimal()+
  theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(add = c(0, 200))) +
  labs(
    title = "What Age groups have the most death?",
    x = "Age groups",
    y = "Count"
  )
```

```{r deaths-age-drugs}
drugs_age <- deaths_drugs %>%
  select(id,age,drug_name,value) %>%
  filter(!is.na(age), drug_name %in% c("heroin","fentanyl","ethanol","cocaine")) %>% 
  mutate(
    age_group = case_when(
      age < 20             ~ "< 20",
      age >= 20 & age < 30 ~ "20-29",
      age >= 30 & age < 40 ~ "30-39",
      age >= 40 & age < 50 ~ "40-49",
      age >= 50 & age < 60 ~ "50-59",
      age >= 60            ~ "> 60",
    ),
    age_group = factor(
      age_group,
      levels = c("< 20", "20-29", "30-39", "40-49", "50-59", "> 60")
    )
  )

drugs_age %>%
  filter(value == 1) %>%
  group_by(age_group, drug_name) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(
    x = age_group,
    y = count,
    fill = age_group,
    label = count
  )) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_text(hjust = -0.2) +
  facet_wrap(vars(drug_name)) +
  theme_minimal()+
  theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(add = c(0, 200))) +
  labs(title = "What age groups have the most death vs weekdays?",
       x = "Age groups",
       y = "Count")
```
## TODO: one more

# Distribution of death in state

It is interesting to analyze where the overdose deaths happened in the state.

Connecticut is composed of 8 counties: Fairfield, Hartford, Litchfield,
Middlesex, New Haven, New London, Tolland, and Windham.
```{r deaths-county}
deaths <- deaths %>%
  mutate(death_county = ifelse(death_county == "Usa", NA, death_county))

deaths_by_county <- deaths %>%
  group_by(death_county) %>%
  filter(!is.na(death_county)) %>%
  summarise(values = n()) %>%
  mutate(death_county = fips(state = "CT", county = death_county)) %>%
  rename(fips = death_county)

plot_usmap(data = deaths_by_county,
           regions = "counties",
           include = "CT") +
  scale_fill_continuous(
    low = "white",
    high = "blue",
    name = "Overdoses per county",
    label = scales::comma
  ) +
  theme(legend.position = "right")
```
Unsurprisingly, most reports are located in the counties of Hartford, Fairfield,
and New Haven, which are the most populated counties in the state.

## Out of state subjects

One interesting statistic is the number of out-of-state individuals, i.e
subjects that live in other US states, but died of overdose in Connecticut.
```{r deaths-out-of-state}
out_of_state <- deaths %>%
  group_by(residence_state) %>%
  filter(!is.na(residence_state),tolower(residence_state)!="ct") %>%
  summarise(values = n()) %>%
  rename(state=residence_state)

sprintf("%0.1f%%", sum(out_of_state[["values"]]) / count(deaths) * 100)
```
The number of out-of-state subject is extremely low.

We can plot the state of residence of such persons:
```{r deaths-out-of-state-map}
plot_usmap(data = out_of_state,
           regions = "state")+
  scale_fill_continuous(
    low = "white",
    high = "blue",
    name = "Overdoses per county",
    label = scales::comma
  ) +
  theme(legend.position = "right")
```
It would seem that most of the people that do not reside in Connecticut
come from the neighboring state of New York.

```{r deaths-city-geo}
geo <- deaths %>%
  group_by(death_city_geo) %>%
  summarise(count=n()) %>%
  mutate(temp=str_extract(death_city_geo, "(?<=\\().+?(?=\\))")) %>%
  mutate(lat=sapply(str_split(temp,","),head,1)) %>%
  mutate(lon=sapply(str_split(temp,","),tail,1)) %>%
  mutate(lat=as.numeric(lat),lon=as.numeric(lon)) %>%
  select(lat,lon,count) 

geo_transformed <- usmap_transform(geo)

plot_usmap(include="CT",regions="counties") +
  geom_point(data = geo_transformed, aes(x=x,y=y,size=count), color = "blue", alpha = 0.35)+ 
   scale_size_continuous(range = c(0.5, 16),
                        labels = scales::comma) +
  labs(title = "Map of deaths by overdose",
       size = "Number of deaths by overdose") +
  theme(legend.position = "right")
```