---
title: "Drug Overdose Analysis"
output:
  pdf_document: 
    keep_tex: yes
  html_notebook: default
documentclass: book
classoption: a4paper
---

# Introduction

In this report we will analyze a dataset concerning the deaths by overdose 
in the U.S. state of Connecticut (CT), between the years 2012 and 2018.

The data set studied was taken at the following link: <https://www.kaggle.com/datasets/ruchi798/drug-overdose-deaths>.
The source of the data set itself is <data.gov>, a website that publishes
freely-accessible information regarding the public administration of the U.S.A.

In chapter 1 the necessary initial steps for the analysis are explained.
In chapter 2, we will continue with the data cleaning operations.
The various topics studied can be found in chapters 3-8.
Finally, in chapter 9, we will summarise all the discoveries
made in the analysis.

## Technical details
This data analysis was made using R v4.1 on RStudio 2022.02.0.
All the code can be found at
<https://github.com/LucaCtt/drug_overdose_analysis>.

# Preparation

## Import packages

The initial step of the analysis is to import the necessary packages:
```{r import-packages}
library("tidyverse")
library("tidytext")
library("janitor")
library("lubridate")
library("usmap")
library("treemap")

theme_set(theme_minimal()) #ggplot2 theme
```
- **tidyverse**: a collection of packages for data science, which includes
  dplyr, ggplot2, and mroe.
- **tidytext**: text mining utilities.
- **janitor**: for examining and cleaning dirty data.
- **lubridate**: utilities for working with dates.
- **usmap**: for US map data frames.
- **treemap**: for treemap plots.

## Import data set


and it contains information regarding accidental drug related deaths in the U.S
state of Connecticut, in the years between 2012-2018.
```{r read-csv}
deaths <- read.csv("drug_deaths.csv")
```
To check that the data was imported correctly, the `head` function can be used
to output the first six records of the data frame.
```{r head-dataset}
head(deaths)
```
The first question that can be asked about the data set is what information it contains,
i.e. what are its columns.
This question can be answered using the `names` function:
```{r names-dataset}
names(deaths)
```
The data set does not include a description for the columns,
but thankfully most of them are self-explaining.
A few, however, require an interpretation:

- *X*: a progressive integer to identify the record, starting from 0.
- *ID*: probably used to identify the person without revealing private information.
        It's composed by a two digit integer that represents the year of death,
        and a progressive four-digit integer that starts from zero.
        The two numbers are separated by a "-".
- *DateType*: specifies if the `date` should be interpreted as the date of death
               or the date when the death was reported.
- *Location*: location of death.
- *LocationifOther*: only used if `location=="other"`, to better specify the
                      location of death.
- *COD*: cause of death.
- *OtherSignificant*: other significant factors related to the death.

Using the functions `nrow` and `ncol` we can get the number of rows and columns
in the data set.
```{r dimensions-dataset}
nrow(deaths)
ncol(deaths)
```

# Data Cleaning

Now that the data has been imported, the data needs to be cleaned
to standardize the records and prevent inconsistencies.

## Names

The `janitor` package provides the `clean_names()` function which handles
problematic variable names, by doing the following cleaning steps:
- Return names with only lowercase letters, with _ as a separator;
- Handle special characters and spaces;
- Append numbers to duplicated names;
- Convert “%” to “percent” to retain meaning.
```{r clean-names}
deaths <- deaths %>% clean_names()
```
The updated names can be seen using the `names` function:
```{r cleaned-names}
names(deaths)
```

## Column modifications

The data set contains two ids for each record: `x` and `id`.
This means that one of them can be removed without consequences.

There could however be multiple records with the same id, which could
be caused by an error in how the data set was built.
This can be verified by computing the number of unique values of `x` in the set,
which should be equal to the number of records.
```{r unique-x}
length(unique(deaths$x)) == nrow(deaths)
```
Since there are no duplicates, the `x` column can be safely deleted:
```{r delete-x}
deaths <- deaths %>% select(-x)
```
Another column that won't be used in this analysis is `DateType`:
```{r delete-date-type}
deaths <- deaths %>% select(-date_type)
```
The data set contains a column named `morphine_not_heroin`.
This is probably due to the fact that heroin is quickly metabolized and 
as such it's hard to detect. Tests to identify heroin usually check
for the presence of morphine, which is a component of heroin.
As such, it's useful to better specify if heroin or morphine were found.

This column can be renamed to just `heroin`.
```{r rename-morphine}
deaths <- deaths %>% rename(morphine = morphine_not_heroin)
```
There also two columns related to fentanyl: `fentanyl` and `fentanyl_analogue`.
This distinction is not useful for this analysis,
so they can be reduced to just one column.
```{r reduce-fentanyl}
deaths <- deaths %>%
  mutate(fentanyl = ifelse(fentanyl_analogue == 1, "1", fentanyl))

deaths <- deaths %>% select(-fentanyl_analogue)
```

## Convert values

There could be empty strings in some cells of the data set.
It's useful to convert such empty strings to NA, so they can be easily
excluded from queries.
```{r mutate-all-na}
deaths <- deaths %>% mutate_all(na_if, "")
```
A few records seem to contain the `Unknown` value for the `sex` field,
which can also be converted to `NA`.
```{r sex-unknown-na}
table(deaths$sex)

deaths <- deaths %>%
  mutate(sex = ifelse(sex == "Unknown", NA, sex))
```
Another couple of columns related to drugs contain weird values:
```{r morphine-fentanyl-weird-values}
deaths %>%
  select(fentanyl, morphine) %>%
  filter((fentanyl != "1" & fentanyl != "0") |
           (morphine != "1" & morphine != "0"))
```
The ones related to fentanyl seem to suggest that the drug was detected,
so they can all be put to `1`.
The cells about morphine instead are unclear so they should be set to NA.
```{r morphine-fentanyl-weird-values-fix}
deaths <- deaths %>%
  mutate(fentanyl = ifelse((fentanyl != "1" &
                              fentanyl != "0"), 1, fentanyl))

deaths <- deaths %>%
  mutate(morphine = ifelse((morphine != "1" &
                              morphine != "0"), NA, morphine))
```

## Inconsistent letter casing

The values in the column `mannerof_death` have inconsistent casing:
```{r inconsistent-casing}
table(deaths$mannerof_death)
```
This problem may also be present in other columns of the data set.
The function `str_to_title` provided by the package `stringr` can be used
to fix this:
```{r inconsistent-casing-fix}
deaths <- deaths %>% mutate_all(str_to_title)
```

## Dates

The data set contains the date of death as a string composed by the
actual date, plus the time of death.

It can be useful to convert the string to an actual date-time class.
However, in the data, for most of the date-times,
the substring that specifies the time is always equal to "12:00:00 AM".
```{r time-always-same}
datetimes <- as.POSIXct(deaths$date, format = "%m/%d/%Y %T")
times <- strftime(datetimes, format = "%H:%M:%S")
print(unique(times))
```
All of the times of death are either "12:00:00 AM" or NA, so they can be eliminated
from the data set.
This means that the significant part of the string is just the date:
```{r set-date}
deaths$date <- as.Date(deaths$date, "%m/%d/%Y")
```

## NA values

We can analyse the percentage of NA values for each column:
```{r plot-na}
missing <- colMeans(is.na(deaths)) * 100
missing <- missing[missing >= 1]
missing <- data.frame(missing)
missing <- rownames_to_column(missing, var = "row_name")

ggplot(data = missing, aes(
  x = fct_reorder(row_name, missing),
  y = missing,
  label = sprintf("%0.1f%%", missing)
)) +
  geom_segment(aes(xend = row_name, y = 0, yend = missing), color = "skyblue") +
  geom_point(color = "blue",
             size = 3,
             alpha = 0.6) +
  geom_text(nudge_y = 7) +
  coord_flip() +
  labs(title = "Percentage of NA values for each column", x = "Column", y =
         "Percentage of NA values")
```
The columns that contain mostly NA values are not particularly
useful, so they can be removed from the data frame.
```{r delete-na-columns}
deaths <- deaths %>%
  select(-c(
    other_significan,
    other,
    locationif_other,
    injury_state,
    injury_county
  ))
```


# Analysis of the abuse of drugs

To simplify the extraction of information about the drug abuse, a new data frame
can be built using `pivot_longer`.

This function allows, for each record (which in the original data frame corresponds
to a single person), to create *n* records in the new data frame, where *n* is
the total number of different drugs reported in the data set (heroin, fentanyl,
morphine, ...). Each of these new *n* records will correspond to a single person,
i.e. they will all have the same `id`.
In each of these new rows, a single drug is associated
to a value of `1` if the drug was reported in the original record,
or `0` otherwise.

In other words, `pivot_longer` lengthens the data, increasing the number of rows
and decreasing the number of columns.
```{r deaths-drugs}
deaths_drugs <- deaths %>%
  select(
    "id",
    "sex",
    "age",
    "location",
    "heroin",
    "fentanyl",
    "morphine",
    "cocaine",
    "oxycodone",
    "oxymorphone",
    "ethanol",
    "hydrocodone",
    "benzodiazepine",
    "methadone",
    "amphet",
    "tramad",
    "hydromorphone"
  ) %>% 
  pivot_longer(
    cols = c(
      "heroin",
      "fentanyl",
      "cocaine",
      "oxycodone",
      "oxymorphone",
      "morphine",
      "ethanol",
      "hydrocodone",
      "benzodiazepine",
      "methadone",
      "amphet",
      "tramad",
      "hydromorphone"
    ),
    names_to = "drug_name",
    values_to = "value",
    values_drop_na = TRUE,
  ) %>%
  mutate(value = as.numeric(value))
```

## Most popular drugs

Using this new data frame, the distribution in the use of
drugs can be easily discovered:
```{r most-popular-drugs}
deaths_drugs %>%
  filter(value != 0) %>%
  group_by(drug_name) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = drug_name,
    y = count,
    label = count,
    fill = drug_name
  )) +
  geom_col(width = 0.5) +
  labs(title = "Drugs found in subjects", x = "Drug",
       y = "Number of subjects") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 0.8)) +
  geom_text(nudge_y = 80)  
```
Heroin and fentanyl, both of which are opioids, are the two most
used drugs by a significant margin.
This could be related to the current opioid epidemic in the USA.

Cocaine is also widespread, along with alcohol consumption.

The fourth place is taken by benzodiazepine,
which is actually a category of drugs that includes many legal medications,
used to treat a variety of disorders. An example of legal benzodiazepine
is Xanax, which is typically used to treat anxiety.

## Number of distinct drugs used

We can also analyze the number of (distinct) drugs used by the subjects.
```{r number-of-drugs}
deaths_drugs %>%
  group_by(id) %>%
  summarise(count = sum(value)) %>%
  count(count) %>%
  ggplot(aes(
    x = count,
    y = n,
    label = n,
    fill = n
  )) +
  geom_col(width = 0.5) +
  labs(title = "Distinct drugs found in subjects",
       x = "Number of drugs", y = "Subjects") +
  theme(legend.position = "none") +
  geom_text(nudge_y = 60)
```

From this plot we can understand that the majority of individuals
overdosed on two distinct drugs at the same.
This is consistent with a study published in 2019, that associates
the combination of drugs to an increased risk of overdose.

The fact that there are subject with no drugs reported is noteworthy:
this could suggest an error in the data insertion, or maybe there were
signs of drug overdose on the patient,
but all the test executed were negative.

## Multiple drugs

We can extract the overall percentage of individuals
that overdosed on multiple drugs:
```{r multiple-drugs}
death_by_multiple_drugs <- deaths_drugs %>%
  group_by(id) %>%
  summarise(count = sum(value)) %>%
  count(count) %>%
  filter(count >= 2) %>%
  reduce(sum)


sprintf("%0.1f%%", death_by_multiple_drugs / count(deaths) * 100)
```

So, the 68.5% of subjects oversdosed on multiple drugs: once again,
this underlines the danger of combining multiple drugs.

## Combination of opioids and benzodiazepines

A particularly lethal combination of drugs are opioids
and benzodiazepines. We can verify how many subjects used these two drugs
at the same time:
```{r opioids-benzodazepines}
death_op_ben <- deaths %>%
  filter(any_opioid == "1", benzodiazepine == "1") %>%
  count()

sprintf("%0.1f%%", death_op_ben / count(deaths) * 100)
```
## Overdoses on single drug

Still, a lot of people overdosed on a single drug. It's interesting to discover
which one it is.
```{r single-drug}
ids <- (deaths_drugs %>%
          group_by(id) %>%
          summarise(count = sum(value)) %>%
          filter(count == 1))[["id"]]

deaths_drugs %>%
  filter(id %in% ids, value == 1) %>%
  group_by(drug_name) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = drug_name,
    y = count,
    label = count,
    fill = drug_name
  )) +
  geom_col(width=0.5) +
  labs(title = "Overdose deaths by single drug", x = "Drug", y = "Number of deaths") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45,vjust=0.8)) +
  geom_text(nudge_y = 20)  
```
So, heroin is the drug that is most overdosed on by itself
with fentanyl coming second.
This is not surprising because, as we have seen before,
these two are the overall most used drugs.

A fascinating result is that ethanol and benzodiazepines, while also
being among the drugs most commonly found in the subjects,
did not cause many overdose victims by themselves.

## Drug usage with alcohol

Another interesting statistic is the number of individuals that used drugs
along with alcohol (ethanol): many studies have confirmed that combining alcohol
with any kind of drug is dangerous.
```{r drugs-with-alcohol}
ids <- (deaths_drugs %>%
          filter(drug_name == "ethanol", value == 1))[["id"]]

used_alcohol <- deaths_drugs %>%
  filter(id %in% ids) %>%
  group_by(id) %>%
  summarise(count = sum(value)) %>%
  count(count) %>%
  filter(count >= 2) %>%
  reduce(sum)

sprintf("%0.1f%%", used_alcohol / count(deaths) * 100)
```
# Analysis of death locations

We can also study the death locations of the subjects.

## Death location distribution

A treemap can be used to show how the locations are distributed.
```{r death-location-distribution}
locations <- deaths %>%
  group_by(location) %>%
  summarise(count = n()) %>%
  select(location,count)

treemap(
  dtf = locations,
  index = "location",
  vSize = "count",
  type = "index",
  
  title = "Death locations",
  
  border.col = c("black"),
  border.lwds = 1,
  
  fontsize.labels = 4,
  fontcolor.labels = "white",
  fontface.labels = 1,
  inflate.labels = TRUE
)
```
The majority of the subjects died at their home, only less than half
of them died in a hospital.

## Death location for most used drugs

It could be interesting to plot the death location for the most used drugs,
to verify if the distribution of locations discovered before is true 
for each drug.
```{r death-location-drug}
deaths_drugs %>%
  mutate(location = fct_lump_n(location, n = 3)) %>%
  select(location, value, drug_name) %>%
  filter(
    location != "Unknown",
    value != 0,
    drug_name %in% c(
      "heroin",
      "fentanyl",
      "cocaine",
      "benzodiazepine",
      "ethanol",
      "oxycodone"
    )
  ) %>%
  group_by(drug_name, location) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  ggplot(aes(
    x = location,
    y = value,
    label = value,
    fill = location
  )) +
  geom_col(width = 0.5) +
  geom_text(vjust = -0.3, size = 4) +
  scale_y_continuous(expand = expansion(add = c(0, 300))) +
  facet_wrap(~ drug_name, scales = "free_x") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45,vjust=0.6)) +
  labs(title = "Death Location by Drug Name",
       x = "Location",
       y = "Count")
```
The results do not hold any surprise.
The most recurring location of death seems to be the subjects resident for
each type of drug.

Only for cocaine, the percentage of people who died in their home is closer
to the percentage of those who died in a hospital.

# Analysis by sex

The data set contains information about both female and male subjects.
To better understand the spread of drug abuse, it is interesting to
discriminate between the two sexes in the analysis.

## Distribution of overdose deaths by sex

We can begin by finding out how many subjects are male and how many are female.
```{r deaths-sex}
deaths %>%
  filter(!is.na(sex), sex != "Unknown") %>%
  group_by(sex) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = sex,
    y = count,
    label = count,
    fill = sex
  )) +
  geom_bar(stat = "identity") +
  labs(title = "Overdose deaths distributed by sex",
       x = "Sex", y = "Number of deaths") +
  geom_text(nudge_y = 200,  size = 5) +
  theme(legend.position = "none")
```

Almost three quarters of all the subjects in the data set are male.
The reason for this is difficult to tell: it could be due to how the data set
was built (i.e. which deaths were included), or it could suggest
that most of the overdose victims are male.

## Most abused drugs for each sex

We can also plot how the abuse of drugs is distributed among the sexes.
```{r drugs-sex}
deaths_drugs %>%
  filter(!is.na(sex), value != 0) %>%
  group_by(drug_name, sex) %>%
  summarise(count = sum(value), .groups = "drop") %>%
  mutate(sex = as.factor(sex),
         drug_name = reorder_within(drug_name, count, sex)) %>%
  ggplot(aes(
    x = fct_reorder(drug_name, count),
    y = count,
    label = count,
    fill = drug_name
  )) +
  geom_col(width = 0.5) +
  scale_y_continuous(expand = expansion(add = c(0, 600))) +
  geom_text(hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_x_reordered() +
  facet_wrap(vars(sex), scales = "free") +
  labs(title = "Most popular drugs for each sex",
       x = "Drug name",
       y = "Number of subjects") +
  theme(legend.position = "none")
```
Heroin and fentanyl are the two most abused drugs by both sexes.
More females than males abused benzodiazepines, which is the third
most used drug by females and the fifth most popular between males.
Also males seem to have a tendency to consume more alcohol than female subjects.

# Analysis by time

An interesting information that can be easily discovered is the distribution
of the deaths over the years.
```{r deaths-years}
deaths %>%
  filter(!is.na(date)) %>%
  group_by(year = format(date, "%Y")) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = year,
             y = count,
             label = count)) +
  geom_segment(aes(xend = year, y = 0, yend = count), color = "skyblue") +
  geom_point(color = "blue",
             size = 3,
             alpha = 0.6) +
  coord_flip() +
  labs(title = "Overdose deaths by year",
       x = "Year", y = "Number of deaths") +
  scale_x_discrete(limits = rev) +
  scale_y_continuous(expand = expansion(add = c(0, 300))) +
  geom_text(nudge_y = 60) 
```
In this plot we can see that the number of overdose is skyrocketing each
following year. Once again this is probably related to the increasing
number of drug users addicted to opiods.
The exception is 2018, where the number deaths is slightly lower than 2017.
This could be because there was an actual decrease in the overdose deaths,
or it could suggest that not all the data for the year was included in the
data set.

## Over the months

To plot the distribution of overdose deaths over the months:
```{r deaths-months}
deaths %>%
  filter(!is.na(date)) %>%
  group_by(month = month(date, label = TRUE)) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = month,
             y = count,
             label = count)) +
  geom_segment(aes(xend = month, y = 0, yend = count), color = "skyblue") +
  geom_point(color = "blue",
             size = 3,
             alpha = 0.6) +
  coord_flip() +
  labs(title = "Overdose deaths by month",
       x = "Month", y = "Number of deaths") +
  scale_x_discrete(limits = rev) +
  scale_y_continuous(expand = expansion(add = c(0, 300))) +
  geom_text(nudge_y = 30)
```
This plot does not contain any particularly useful information:
the number of overdose cases is quite consistent over the months of the years.
The slight fluctuations could be caused by the availability of drugs.

## Over the days of week

Finally, we can plot the number of deaths for each day of the week.
```{r deaths-day}
deaths %>%
  filter(!is.na(date)) %>%
  group_by(day = wday(date, label = TRUE)) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = day,
    y = count,
    label = count
  )) +
  geom_segment(aes(xend = day, y = 0, yend = count), color = "skyblue") +
  geom_point(color = "blue",
             size = 3,
             alpha = 0.6) +
  coord_flip() +
  labs(title = "Overdose deaths by day of week",
       x = "Day of week", y = "Number of deaths") +
  scale_x_discrete(limits = rev) +
  scale_y_continuous(expand = expansion(add = c(0, 300))) +
  geom_text(nudge_y = 40)
```
The number of deaths is higher during the weekends: this could be caused
by parties and increased nightlife, where it is easier to find drugs.

# Analysis by age

It's also interesting to study the age of the subjects.
For this purpose, we can create a new data frame that
contains, for each subject, the age group they belong to.

The age groups considered are:
- < 20
- 20-29
- 30-39
- 40-49
- 50-59
- >= 60

```{r age-groups}
deaths_age <- deaths %>%
  select(id,age) %>%
  filter(!is.na(age)) %>% 
  mutate(
    age_group = case_when(
      age < 20             ~ "< 20",
      age >= 20 & age < 30 ~ "20-29",
      age >= 30 & age < 40 ~ "30-39",
      age >= 40 & age < 50 ~ "40-49",
      age >= 50 & age < 60 ~ "50-59",
      age >= 60            ~ "> 60",
    ),
    age_group = factor(
      age_group,
      levels = c("< 20", "20-29", "30-39", "40-49", "50-59", ">= 60")
    )
  )
```

## Distribution of deaths by age group

```{r deaths-age}
deaths_age %>%
  group_by(age_group) %>%
  summarise(count = n()) %>%
  ggplot(aes(
    x = age_group,
    y = count,
    fill = age_group,
    label = count
  )) +
  geom_col(width = 0.5) +
  coord_flip() +
  geom_text(hjust = -0.2) +
  theme(legend.position = "none") +
  scale_x_discrete(limits = rev) +
  scale_y_continuous(expand = expansion(add = c(0, 500))) +
  labs(
    title = "Distribution of deaths by age group",
    x = "Age group",
    y = "Number of subjects"
  )
```

```{r deaths-age-drugs}
drugs_age <- deaths_drugs %>%
  select(id,age,sex,drug_name,value) %>%
  filter(!is.na(age), drug_name %in% c("heroin","fentanyl","ethanol","cocaine")) %>% 
  mutate(
    age_group = case_when(
      age < 20             ~ "< 20",
      age >= 20 & age < 30 ~ "20-29",
      age >= 30 & age < 40 ~ "30-39",
      age >= 40 & age < 50 ~ "40-49",
      age >= 50 & age < 60 ~ "50-59",
      age >= 60            ~ "> 60",
    ),
    age_group = factor(
      age_group,
      levels = c("< 20", "20-29", "30-39", "40-49", "50-59", "> 60")
    )
  )

drugs_age %>%
  filter(value == 1) %>%
  group_by(age_group, drug_name) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(
    x = age_group,
    y = count,
    fill = age_group,
    label = count
  )) +
  geom_col(width = 0.5) +
  coord_flip() +
  geom_text(hjust = -0.2) +
  facet_wrap(vars(drug_name)) +
  theme(legend.position = "none") +
  scale_x_discrete(limits = rev) +
  scale_y_continuous(expand = expansion(add = c(0, 200))) +
  labs(title = "Distribution of drug used by age group",
       x = "Age group",
       y = "Number of subjects")
```
The plots show that cocaine is more popular in the 40-49 age group,
while ethanol is used most by the 50-59 age group.
Fentanyl and heroin are preferred by the 30-39 age group.

## Age and sex

```{r age-sex}
drugs_age %>%
  filter(!is.na(sex), value != 0) %>%
  group_by(age_group, sex) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(sex = as.factor(sex)) %>%
  ggplot(aes(
    x = age_group,
    y = count,
    label = count,
    fill = age_group
  )) +
  geom_col(width = 0.4) +
  scale_x_discrete(limits = rev) +
  scale_y_continuous(expand = expansion(add = c(0, 600))) +
  geom_text(hjust = -0.2, size = 3.5) +
  coord_flip() +
  facet_wrap(vars(sex), scales = "free") +
  labs(title = "Distribution of sexes in age groups",
       x = "Age group",
       y = "Number of subjects") +
  theme(legend.position = "none")
```
Most of the females belong in the 40-49 age group, while most of the males
in the died at age 30-39.

# Analysis by geographical location

It is interesting to analyze the geographical location of overdose deaths.

## Analysis by county

Connecticut is composed of 8 counties: Fairfield, Hartford, Litchfield,
Middlesex, New Haven, New London, Tolland, and Windham.
```{r deaths-county}
deaths <- deaths %>%
  mutate(death_county = ifelse(death_county == "Usa", NA, death_county))

deaths_by_county <- deaths %>%
  group_by(death_county) %>%
  filter(!is.na(death_county)) %>%
  summarise(values = n()) %>%
  mutate(death_county = fips(state = "CT", county = death_county)) %>%
  rename(fips = death_county)

plot_usmap(data = deaths_by_county,
           regions = "counties",
           labels=TRUE,
           include = "CT") +
  scale_fill_continuous(
    na.value="white",
    low = "lightcyan",
    high = "dodgerblue4",
    name = "Overdoses per county",
    label = scales::comma
  ) +
  theme(legend.position = "right")
```
Unsurprisingly, most reports are located in the counties of Hartford, Fairfield,
and New Haven, which are the most populated counties in the state.

## Out of state subjects

One interesting statistic is the number of out-of-state individuals, i.e
subjects that live in other US states, but died of overdose in Connecticut.
```{r deaths-out-of-state}
out_of_state <- deaths %>%
  group_by(residence_state) %>%
  filter(!is.na(residence_state),tolower(residence_state)!="ct") %>%
  summarise(values = n()) %>%
  rename(state=residence_state)

sprintf("%0.1f%%", sum(out_of_state[["values"]]) / count(deaths) * 100)
```
The number of out-of-state subject is extremely low.

We can nevertheless plot the state of residence of such persons:
```{r deaths-out-of-state-map}
plot_usmap(data = out_of_state,
           regions = "state")+
  scale_fill_continuous(
    na.value="white",
    low = "lightcyan",
    high = "dodgerblue4",
    name = "Number of residents that overdosed in CT",
    label = scales::comma
  ) +
  theme(legend.position = "right")
```
It would seem that most of the people that do not reside in Connecticut
come from the neighboring states of New York and Massachussets.

## Analysis by city of death

```{r deaths-city-geo}
geo <- deaths %>%
  group_by(death_city_geo) %>%
  summarise(count = n()) %>%
  mutate(temp = str_extract(death_city_geo, "(?<=\\().+?(?=\\))")) %>%
  mutate(lat = sapply(str_split(temp, ","), head, 1)) %>%
  mutate(lon = sapply(str_split(temp, ","), tail, 1)) %>%
  mutate(lat = as.numeric(lat), lon = as.numeric(lon)) %>%
  select(lat, lon, count)

geo_transformed <- usmap_transform(geo)

plot_usmap(include = "CT", regions = "counties") +
  geom_point(
    data = geo_transformed,
    aes(x = x, y = y, size = count),
    color = "dodgerblue4",
    alpha = 0.6
  ) +
  scale_size_continuous(range = c(0.5, 16),
                        labels = scales::comma) +
  labs(title = "Map of deaths by overdose",
       size = "Number of deaths by overdose") +
  theme(legend.position = "right")
```
As expected, most deaths happened in the big cities of the state.